Elaboration with First-Class Implicit Function Types

### 2020-02-23 paper.tex
stuff

\newcommand{\slet}{\boldsymbol{\mathsf{let}}}
\renewcommand{\sin}{\boldsymbol{\mathsf{in}}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\To}{\Rightarrow}
\newcommand{\p}{\mathsf{p}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\freshMeta}[3]{\mathsf{freshMeta}\,_{#1|#2}\,#3}
\newcommand{\constancy}{\mathsf{constancy}}
\newcommand{\newConstancy}[6]{\mathsf{newConstancy}_{#1|#2,\,#3 : \Rec\,(#4\,#5)}\,#6}
\newcommand{\unify}{\mathsf{unify}}
\newcommand{\fail}{\mathsf{fail}}
\newcommand{\kw}[1]{{\mathsf{#1}}}
\newcommand{\FreeVars}{\mathsf{FreeVars}}

@@

\section{Basic Bidirectional Elaboration}
\label{sec:basic_bidirectional_elaboration}

First, we present a variant of Norell's bidirectional elaborator
\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
simplifications; what we end up with can be viewed as a toy version of the
actual Agda elaborator. In this section, we use it to illustrate the key issues,
and we extend it in Section \ref{sec:extending_elaboration}.
\begin{figure}[h]
\begin{alignat*}{4}
  t,\,u,\,v,\,A,\,B,\,C\, :&:=\quad  && x\hspace{8em}              & \text{variable}                 &  \\
                           & |       && (x : A)\to B               & \text{function type}            &  \\
                           & |       && \{x : A\}\to B             & \text{implicit function type}   &  \\
                           & |       && t\,u                       & \text{application}              &  \\
                           & |       && t\,\{u\}                   & \text{implicit application}     &  \\
                           & |       && \lambda\,x.\, t            & \text{lambda abstraction}       &  \\
                           & |       && \lambda\,\{x\}.\,t         & \text{implicit abstraction}     &  \\
                           & |       && \U                         & \text{universe}                 &  \\
                           & |       && \slet\,x : A = t\,\sin\, u & \text{let-definition}           &  \\
                           & |       && \_                         & \text{hole for inferred term}   &
\end{alignat*}
\caption{Syntax of the surface language.}
\label{fig:surface}
\end{figure}

@@

\begin{notation}\label{not:spines}
  We sometimes use a spine notation for neutral terms. A spine is a list of
  terms, noted as $\overline{t}$, where terms may be wrapped in brackets to
  signal implicit application. For example, if $\overline{u} \equiv
  (\{A\},\,\{B\},\,x)$, then $t\,\overline{u}$ denotes $t\,\{A\}\,\{B\}\,x$.  In
  $t\,\overline{u}$, we call $t$ the \emph{head} of the neutral term. In
  particular, if $t$ is a metavariable, the neutral term is \emph{meta-headed}.
\end{notation}

\begin{example}
The core syntax is quite expressive as a programming language, thanks to
$\slet$-definitions\footnote{In dependent type theories, the $\slet$ rule is not
  derivable from function application, unlike in simple type theories.} and the
type-in-type rule which allows Church-encodings of a large class of inductive
types. For example, the following term computes a list of types by mapping:
\begin{alignat*}{3}
  & \slet\,List : \U\to\U\\
  & \hspace{1em}= \lambda\,A.\,(L : \U)\to(A\to L\to L)\to L\to L\,\,\sin\\
  & \slet\,map : \{A\,B : \U\}\to (A \to B) \to List\,A \to List\,B\\
  & \hspace{1em}=
  \lambda\,\{A\}\,\{B\}\,f\,as\,L\,cons\,nil.\,as\,L\,(\lambda\,a.\,cons\,(f\,a))\,nil\,\,\sin\\
  & map\,\{\U\}\,\{\U\}\,(\lambda\,A.\, A \to A)\,(\lambda\,L\,cons\,nil.\,cons\,\U\,(cons\,\U\,nil))
\end{alignat*}
In the core syntax, all applications must be explicit. In the surface syntax we
will be able to omit implicit arguments to $map$.

\end{example}

@@

\begin{itemize}
  \item
    A metasubstitution $\boxed{\theta : \Theta_0 \To \Theta_1}$ assigns to each variable
    in $\Theta_1$ a term in $\Theta_0$ , hence it is represented as a list of
    terms $(\alpha_1 \mapsto t_1,\,...\,\,\alpha_i \mapsto t_i)$.
  \item
    We define the action of a metasubstitution on contexts and terms by
    recursion; we notate action on contexts as $\Gamma[\theta]$ and action on
    terms as $t[\theta]$. We remark that there is no abstraction for
    metavariables in the core syntax, so we do not have to handle variable
    capture (or index shifting).
\end{itemize}
The following are admissible:
\begin{mathpar}
  \inferrule*[lab=metasub/empty]
             {\Theta_0\vdash}
             {() : \Theta_0\To \emptycon}

@@

  \inferrule*[lab=metasub/con-action]
             {\Theta_1|\Gamma\vdash \\ \theta : \Theta_0\To\Theta_1}
             {\Theta_0|\Gamma[\theta]\vdash}

  \inferrule*[lab=metasub/tm-action]
             {\Theta_1|\Gamma\vdash t : A \\ \theta : \Theta_0\To\Theta_1}
             {\Theta_0|\Gamma[\theta] \vdash t[\theta] : A[\theta]}

@@

\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in implicit
    argument insertion and unification.
\end{itemize}

@@

                \echeck{u}{\Theta_2}{\Gamma[\theta_1\circ\theta_2]}{(B[\theta_1\circ\theta_2])}\\
  &\quad\ereturn\,(\Theta_3,\,\theta_1\circ\theta_2\circ\theta_3,\,\slet\,x:A'[\theta_2\circ\theta_3]=t'[\theta_3]\,\sin\,u')\\

@@

  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,\,t'[\theta_2])\\\\

@@

  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,

@@

  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,\,\lambda\,x.\,t',\,(x : A[\theta_2])\to B)\\

@@

  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,x:A_0}{\U}\\
  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_2\circ\theta_3])\,((x : A_0[\theta_3])\to A_1)\\
  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4]}{(A_0[\theta_3\circ\theta_4])}\\
  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_2\circ\theta_3\circ\theta_4\circ\theta_5])\,u')}\,{(A_1[\theta_4\circ\theta_5][x\mapsto u'])})\\

@@

  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,x:A_0}{\U}\\
  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_2\circ\theta_3])\,(\{x : A_0[\theta_3]\}\to A_1)\\
  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4]}{(A_0[\theta_3\circ\theta_4])}\\
  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_2\circ\theta_3\circ\theta_4\circ\theta_5])\,\{u'\})}\,{(A_1[\theta_4\circ\theta_5][x\mapsto u'])})\\

@@

                \einfer{u}{ins}{\Theta_2}{\Gamma[\theta_1\circ\theta_2]}\\
  &\quad\ereturn\,(\Theta_3,\,\theta_1\circ\theta_2\circ\theta_3,\,(\slet\,x:A'[\theta_2\circ\theta_3]=t'[\theta_3]\,\sin\,u'),\,B)\\

@@

\begin{itemize}
  \item \emph{Let-generalization}. This is an open research topic in settings
    with dependent types, and we make no attempt at covering it. See
    \cite{eisenberg2016dependent} for a treatment in a proposed dependent
    version of Haskell.
  \item \emph{Polymorphic subtyping}. In some prior works, e.g.\ in
    \cite{dunfield2013complete, vytiniotis2008fph}, there is a subtyping
    relation arising along instantiations of polymorphic types. In GHC 8
    polymorphic subtyping is implemented for function types only. Polymorphic
    subtyping complicates type inference, and to our knowledge it has not been
    implemented in any dependently typed setting. We also believe that it is an
    anti-feature in dependent settings, because elaboration of subtyping must
    insert coercions which significantly change the intensional character of
    programs. For example, if we have covariant list types, then coercing $t :
    List\,(\{A : \U\}\to A \to A)$ to $t : List\,(Bool \to Bool)$ requires
    mapping over $t$ and inserting implicit applications to $Bool$ for each list
    element. In System F, all such coercions are erasible, since types are
    computationally irrelevant, but in our core syntax we have implicit
    functions with arbitrary (relevant) domains. In GHC, subtyping coercions for
    functions change operational semantics; this is a reason for abandoning
    subtyping in recent developments of impredicative inference for GHC
    \cite{serrano2020a}.
\end{itemize}

@@

\begin{definition}[Closing types]
We use curried function space to close over record types in the scope. We
prepend the following clause to Definition \ref{def:closingtype}:
\[
  ((\Gamma,\, x : \Rec\,A) \To B) :\equiv (\Gamma \To (\{x : \ol{A}\} \to B))
\]
\end{definition}
\begin{definition}[Contextualization]
We extend our spine notation to applications of curried functions. For example,
we may have a spine $\ol{t} \equiv (\{x : \ol{A}\}\,\{y : \ol{B}\})$. We
accordingly revise Definition \ref{def:contextualization} for
$\overline{\mathsf{vars}_{\Gamma}}$ so that we use curried function application
for each record type in $\Gamma$.
\end{definition}

@@

\subsubsection{Constancy constraints}
We use \emph{constancy constraints} to get rid of curried function types as
soon as we learn that they are non-dependent. These are constraints in the usual
sense in unification algorithms (e.g.\ as in \cite{abel2011higher} or
\cite{vytiniotis2011outsidein}). We formalize them in a compact way, by adding a
new kind of context extension for metacontexts. The rules are given in Figure
\ref{fig:constancy}.

\begin{figure}
\begin{mathpar}
  \inferrule*[lab=metacon/constancy]
             {\Theta|\emptycon \vdash \alpha : \Delta \To \Tel \\ \Theta|\Gamma,\,x : \Rec\,(\alpha\,\ol{t}) \vdash B : \U}
             {\Theta,\,\mathsf{constancy}_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B\vdash}

  \inferrule*[lab=constancy-$\equiv$]
             {\Theta_0|\emptycon \vdash \alpha : \Delta \To \Tel \\ x \notin \FreeVars(B)}
             {\Theta_0,\,\mathsf{constancy}_{\Gamma_0,\,x : \Rec\,(\alpha\,\ol{t})}\,B,\,\Theta_1|\Gamma_1
               \vdash \alpha \equiv (\lambda\,\overline{\mathsf{vars}_{\Delta}}.\,\epsilon) : \Delta \To \Tel}

  \inferrule*[lab=metasub/constancy]
             {\Theta_1|\emptycon \vdash \alpha : \Delta \To \Tel \\ \theta : \Theta_0 \To \Theta_1 \\\\ (x \notin \FreeVars(B[\theta]))\,\,\,\text{implies}\,\,\,(\alpha[\theta] \equiv (\lambda\,\overline{\mathsf{vars}_{\Delta[\theta]}}.\,\epsilon))}
             {(\theta,\,\mathsf{solve}) : \Theta_0 \To (\Theta_1,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B)}

  \inferrule*[lab=metasub/weaken-constancy]
             {\\}
             {\p : (\Theta,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B) \To \Theta}
\end{mathpar}
\caption{Rules for constancy constraints}
\label{fig:constancy}
\end{figure}

In the rule \LabTirName{metacon/constancy} we specify extension of a
metacontext with a constraint. The metavariable $\alpha : \Gamma \To \Tel$
represents an unknown domain for a curried function type, and $B$ is the
codomain of the function.

The \LabTirName{constancy-$\equiv$} rule expresses that, assuming we have a
constancy constraint for $\alpha$ and $B$ in context, if $B$ does not depend on
the $x : \Rec\,(\alpha\,\ol{t})$ domain variable, then $\alpha$ is equal to the
constant function which returns the empty telescope. We use the abbreviation
$\lambda\,\overline{\mathsf{vars}_{\Gamma_0}}$ to refer to an iterated $\lambda$
which abstracts over all variables in $\Gamma_0$.

The \LabTirName{metasub/constancy} rule defines metasubstitutions whose
codomains are extended with constraints. Intuitively, while the
\LabTirName{metasub/extended} rule from Section \ref{sec:metasubstitutions} can
be used to solve a metavariable (by mapping it to a term),
\LabTirName{metasub/constancy} solves a constraint. We can only extend $\theta :
\Theta_0 \To \Theta_1$ to map into an additional constraint in the codomain if
$\theta$ forces the constraint to hold. In
\LabTirName{metasub/weaken-constancy}, we overload $\p$ for the weakening
substitution which drops a constraint.

\begin{definition}[Creating a new constraint]
We define this similarly to Definition \ref{def:freshmeta}, by simply
returning a weakening substitution.
\[
\newConstancy{\Theta}{\Gamma}{x}{\alpha}{\ol{t}}{B} : \equiv
             ((\Theta,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B),\,\p)
\]
\end{definition}

@@

After solving a meta $\beta$ during unification, which yields a unifying
$\theta$ substitution, we review all ($\constancy_{\Gamma,\,x :
  \Rec\,(\alpha\,\ol{t})}\,B$) constraints in the context, where $x$ occurs in
$B$ inside a $\ol{t}$ spine of some $\beta\,\ol{t}$ term. In other words, we
review constraints where the new meta solution might make a difference.
\begin{enumerate}
  \item If we have $x \in \FreeVars(B[\theta])$, where $x$ occurs rigidly in
    $B[\theta]$, i.e.\ the occurrence is not in a spine of a meta, then no
    metasubstitution can possibly remove this occurrence. In this case the
    constraint holds vacuously, so we can use the rule
    \LabTirName{metasub/constancy} to return a $\theta'$ substitution which also
    solves the constraint.
  \item If we have $x \notin \FreeVars(B[\theta])$, we solve $\alpha$ to be
    constantly $\epsilon$, thereby getting a $\theta'$ substitution. Since
    $\theta'$ forces the constrain to hold, we can again use
    \LabTirName{metasub/constancy} to solve the constraint. Then, since we have
    just solved $\alpha$, we recursively check for solvable constraints.
  \item
    In any other case we simply return $\theta$ and keep the constraint around.
\end{enumerate}

\emph{Remark.} In the case with $x \in \FreeVars(B[\theta])$, it would be also
sound to solve the constraint when the occurrence is not necessarily
rigid. However, this way we could lose potential non-$\epsilon$ solutions of
$\alpha$.

\subsection{Unification For Strictly Curried Functions}
Although we omit most details of unification, we shall discuss it for curried
functions, as it is essential in the extended elaboration algorithm.  The most
interesting case is when we unify a curried function type with an implicit
function type. In this case, we learn that the domain of the curried function is
non-empty, so we refine the domain telescope.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x : A'\}\to B') :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1,\,\alpha\,\ol{t}) \leftarrow \freshMeta{\Theta_0}{\Gamma,\,x : A'}{\Tel}\\
& \quad (\Theta_2,\,\theta_2) \leftarrow
  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A'[\theta_1])\TCons \alpha\,\ol{t})\\
& \quad (\Theta_3,\,\theta_3) \leftarrow \newConstancy{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,y : A'[\theta_1\circ\theta_2]}{z}{(\alpha}{\ol{t})[\theta_2]}{(B[\theta_1\circ\theta_2][x\mapsto (y :: z)])}\\
& \unify_{\Theta_3|\Gamma[\theta_{123}]}\,(\{z : \ol{\alpha\,\ol{t}}
\end{alignat*}

>>>>

\newcommand{\slet}{\boldsymbol{\mathsf{let}}}
\renewcommand{\sin}{\boldsymbol{\mathsf{in}}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\To}{\Rightarrow}
\newcommand{\p}{\mathsf{p}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\freshMeta}[3]{\mathsf{freshMeta}\,_{#1|#2}\,#3}
\newcommand{\constancy}{\mathsf{constancy}}
\newcommand{\newConstancy}[5]{\mathsf{newConstancy}_{#1|#2,\,#3 : \Rec\,#4}\,#5}
\newcommand{\unify}{\mathsf{unify}}
\newcommand{\fail}{\mathsf{fail}}
\newcommand{\kw}[1]{{\mathsf{#1}}}
\newcommand{\FreeVars}{\mathsf{FreeVars}}

@@

\section{Bidirectional Elaboration}
\label{sec:bidirectional_elaboration}

First, we present a variant of Norell's bidirectional elaborator
\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
simplifications; what we end up with can be viewed as a toy version of the
actual Agda elaborator. In this section, we use it to build the backbone of our
algorithm and illustrate the key issues. We extend this elaborator in Section
\ref{sec:extending_elaboration}.
\begin{figure}[h]
\begin{alignat*}{4}
  t,\,u,\,v,\,A,\,B,\,C\, :&:=\quad  && x\hspace{8em}              & \text{variable}                 &  \\
                           & |       && (x : A)\to B               & \text{function type}            &  \\
                           & |       && \{x : A\}\to B             & \text{implicit function type}   &  \\
                           & |       && t\,u                       & \text{application}              &  \\
                           & |       && t\,\{u\}                   & \text{implicit application}     &  \\
                           & |       && \lambda\,x.\, t            & \text{lambda abstraction}       &  \\
                           & |       && \lambda\,\{x\}.\,t         & \text{implicit abstraction}     &  \\
                           & |       && \U                         & \text{universe}                 &  \\
                           & |       && \slet\,x : A = t\,\sin\, u & \text{let-definition}           &  \\
                           & |       && \_                         & \text{hole for inferred term}   &
\end{alignat*}
\caption{Syntax of the surface language.}
\label{fig:surface}
\end{figure}

@@

\begin{notation}\label{not:spines}
  We use a spine notation for neutral terms. A spine is a list of
  terms, noted as $\overline{t}$, where terms may be wrapped in brackets to
  signal implicit application. For example, if $\overline{u} \equiv
  (\{A\},\,\{B\},\,x)$, then $t\,\overline{u}$ denotes $t\,\{A\}\,\{B\}\,x$.  In
  $t\,\overline{u}$, we call $t$ the \emph{head} of the neutral term. In
  particular, if $t$ is a metavariable, the neutral term is \emph{meta-headed}.
\end{notation}

\begin{example}
The core syntax is quite expressive as a programming language, thanks to
$\slet$-definitions\footnote{In dependent type theories, the $\slet$ rule is not
  derivable from function application, unlike in simple type theories.} and the
type-in-type rule which allows Church-encodings of a large class of inductive
types. For example, the following term computes a list of types by mapping:
\begin{alignat*}{3}
  & \slet\,List : \U\to\U\\
  & \hspace{1em}= \lambda\,A.\,(L : \U)\to(A\to L\to L)\to L\to L\,\,\sin\\
  & \slet\,map : \{A\,B : \U\}\to (A \to B) \to List\,A \to List\,B\\
  & \hspace{1em}=
  \lambda\,\{A\}\,\{B\}\,f\,as\,L\,cons\,nil.\,as\,L\,(\lambda\,a.\,cons\,(f\,a))\,nil\,\,\sin\\
  & map\,\{\U\}\,\{\U\}\,(\lambda\,A.\, A \to A)\,(\lambda\,L\,cons\,nil.\,cons\,\U\,(cons\,\U\,nil))
\end{alignat*}
\end{example}
%% In the core syntax, all abstractions and applications must be notated. Thanks to
%% elaboration, in the surface syntax we may omit some of these.

@@

\begin{itemize}
  \item
    A metasubstitution $\boxed{\theta : \Theta_0 \To \Theta_1}$ assigns to each variable
    in $\Theta_1$ a term in $\Theta_0$ , hence it is represented as a list of
    terms $(\alpha_1 \mapsto t_1,\,...\,\,\alpha_i \mapsto t_i)$.
  \item
    We define the action of a metasubstitution on contexts and terms by
    recursion; we notate action on contexts as $\Gamma[\theta]$ and action on
    terms as $t[\theta]$. We remark that there is no abstraction for
    metavariables in the core syntax, so we do not have to handle variable
    capture (or index shifting).
\end{itemize}
The following are admissible:
\begin{mathpar}
  \inferrule*[lab=metasub/empty]
             {\Theta\vdash}
             {() : \Theta\To \emptycon}

@@

  \inferrule*[lab=metasub/con-action]
             {\theta : \Theta_0\To\Theta_1 \\ \Theta_1|\Gamma\vdash}
             {\Theta_0|\Gamma[\theta]\vdash}

  \inferrule*[lab=metasub/tm-action]
             {\theta : \Theta_0\To\Theta_1 \\ \Theta_1|\Gamma\vdash t : A}
             {\Theta_0|\Gamma[\theta] \vdash t[\theta] : A[\theta]}

@@

\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in implicit
    argument insertion and unification.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$, $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}

@@

                \echeck{u}{\Theta_2}{\Gamma[\theta_{12}]}{(B[\theta_{12}])}\\
  &\quad\ereturn\,(\Theta_3,\,\theta_{123},\,\slet\,x:A'[\theta_{23}]=t'[\theta_3]\,\sin\,u')\\

@@

  &\quad\ereturn\,(\Theta_2,\,\theta_{12},\,t'[\theta_2])\\\\

@@

  &\quad\ereturn\,(\Theta_2,\,\theta_{12},

@@

  &\quad\ereturn\,(\Theta_2,\,\theta_{12},\,\lambda\,x.\,t',\,(x : A[\theta_2])\to B)\\

@@

  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\
  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_{23}])\,((x : A_0[\theta_3])\to A_1)\\
  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_{1234}]}{(A_0[\theta_{34}])}\\
  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_{2345}])\,u')}\,{(A_1[\theta_{45}][x\mapsto u'])})\\

@@

  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\
  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_{23}])\,(\{x : A_0[\theta_3]\}\to A_1)\\
  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_{1234}]}{(A_0[\theta_{34}])}\\
  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_{2345}])\,\{u'\})}\,{(A_1[\theta_{45}][x\mapsto u'])})\\

@@

                \einfer{u}{ins}{\Theta_2}{\Gamma[\theta_{12}]}\\
  &\quad\ereturn\,(\Theta_3,\,\theta_{123},\,(\slet\,x:A'[\theta_{23}]=t'[\theta_3]\,\sin\,u'),\,B)\\

@@

\begin{itemize}
  \item \emph{Let-generalization}. This is an open research topic in settings
    with dependent types, and we make no attempt at covering it. See
    \cite{eisenberg2016dependent} for a treatment in a proposed dependent
    version of Haskell.
  \item \emph{Polymorphic subtyping}. In some prior works, e.g.\ in
    \cite{dunfield2013complete, vytiniotis2008fph}, there is a subtyping
    relation arising along instantiations of polymorphic types. In GHC 8
    polymorphic subtyping is implemented for function types only. Polymorphic
    subtyping complicates type inference, and to our knowledge it has not been
    implemented in any dependently typed setting. We also believe that it is
    undesirable in dependent settings, because elaboration of subtyping must
    insert coercions which significantly change the intensional character of
    programs. For example, if we have covariant list types, then coercing $t :
    List\,(\{A : \U\}\to A \to A)$ to $t : List\,(Bool \to Bool)$ requires
    mapping over $t$ and inserting implicit applications to $Bool$ for each list
    element. In System F, all such coercions are erasible, since types are
    computationally irrelevant, but in our core syntax we have implicit
    functions with arbitrary (relevant) domains. In GHC, subtyping coercions for
    functions change operational semantics; this is a reason for abandoning
    subtyping in recent developments of impredicative inference for GHC
    \cite{serrano2020a}.
\end{itemize}

@@

\begin{definition}[Closing types]
We use curried function types to close over record types in the scope. If a
bound variable does not have a record type, then we do as before\footnote{This
  implies that we close over meta-headed types using plain functions. In theory,
  this causes a higher-order version of the basic implicit insertion problem: we
  are uncertain about whether we should be uncertain about implicit
  insertions. So far, this higher-order insertion problem seems to seems to be
  irrelevant in practice, in the prototype implementation.}. We prepend the
following clause to Definition
\ref{def:closingtype}:
\[
  ((\Gamma,\, x : \Rec\,A) \To B) :\equiv (\Gamma \To (\{x : \ol{A}\} \to B))
\]
\end{definition}
\begin{definition}[Contextualization]
We extend our spine notation to applications of curried functions. For example,
we may have a spine $\ol{t} \equiv (\{x : \ol{A}\}\,\{y : \ol{B}\})$. We
accordingly revise Definition \ref{def:contextualization} for
$\overline{\mathsf{vars}_{\Gamma}}$ so that we use curried function application
for each record type in $\Gamma$.
\end{definition}

@@

\subsubsection{Constancy constraints}
We use these constraints to get rid of curried function types as soon as we
learn that they are non-dependent. They are constraints in the usual sense in
unification algorithms (e.g.\ as in \cite{abel2011higher} or
\cite{vytiniotis2011outsidein}). We formalize them in a compact way, by adding a
new kind of context extension for metacontexts. The rules are given in Figure
\ref{fig:constancy}.

\begin{figure}
\begin{mathpar}
  \inferrule*[lab=metacon/constancy]
             {\Theta|\Gamma,\,x : \Rec\,A \vdash B : \U}
             {\Theta,\,\mathsf{constancy}_{\Gamma,\,x : \Rec\,A}\,B\vdash}

  \inferrule*[lab=constancy-$\equiv$]
             {x \notin \FreeVars(B)}
             {\Theta_0,\,\mathsf{constancy}_{\Gamma_0,\,x : \Rec\,A}\,B,\,\Theta_1|\Gamma_1 \vdash A \equiv \epsilon : \Tel}

  \inferrule*[lab=metasub/constancy]
             {\theta : \Theta_0 \To \Theta_1 \\\\ x \notin \FreeVars(B[\theta])\,\,\,\text{implies}\,\,\,\Theta_0|\Gamma[\theta]\vdash A[\theta] \equiv \epsilon : \Tel}
             {(\theta,\,\mathsf{solve}_{\Gamma,\,x : \Rec\,A}\,B) : \Theta_0 \To (\Theta_1,\,\constancy_{\Gamma,\,x : \Rec\,A}\,B)}

  \inferrule*[lab=metasub/weaken-constancy]
             {\\}
             {\p : (\Theta,\,\constancy_{\Gamma,\,x : \Rec\,A}\,B) \To \Theta}
\end{mathpar}
\caption{Rules for constancy constraints}
\label{fig:constancy}
\end{figure}

In the rule \LabTirName{metacon/constancy} we specify extension of a metacontext
with a constraint. The \LabTirName{constancy-$\equiv$} rule expresses that,
assuming we have a constancy constraint for $A$ and $B$ in context, if $B$ does
not depend on the $x : \Rec\,A$ domain variable, then $A$ is equal to the the
empty telescope $\epsilon$.

The \LabTirName{metasub/constancy} rule defines metasubstitutions whose
codomains are extended with constraints. Intuitively, while the
\LabTirName{metasub/extended} rule from Section \ref{sec:metasubstitutions} can
be used to solve a metavariable (by mapping it to a term),
\LabTirName{metasub/constancy} solves a constraint. We can only extend $\theta :
\Theta_0 \To \Theta_1$ to map into an additional constraint if $\theta$ forces
the constraint to hold. In \LabTirName{metasub/weaken-constancy}, we overload
$\p$ for the weakening substitution which drops a constraint.

\begin{definition}[Creating a new constraint]
We do this similarly to Definition \ref{def:freshmeta}, by simply
returning a weakening substitution.
\[
\newConstancy{\Theta}{\Gamma}{x}{A}{B} : \equiv
             ((\Theta,\,\constancy_{\Gamma,\,x : \Rec\,A}\,B),\,\p)
\]
\end{definition}

@@

After solving a meta $\alpha$ during unification, which yields a unifying
$\theta$ substitution, we review all ($\constancy_{\Gamma,\,x : \Rec\,A}\,B$)
constraints in the context, such that $x$ occurs in $B$ inside a $\ol{t}$ spine
of some $\alpha\,\ol{t}$ term. In other words, we review constraints where the
new meta solution might make a difference.
\begin{enumerate}
  \item If we have $x \in \FreeVars(B[\theta])$, where $x$ occurs rigidly in
    $B[\theta]$, i.e.\ the occurrence is not in a spine of a meta, then no
    metasubstitution can possibly remove this occurrence. In this case the
    constraint holds vacuously, so we can use the rule
    \LabTirName{metasub/constancy} to return a $\theta'$ substitution which also
    solves the constraint.
  \item If we have $x \notin \FreeVars(B[\theta])$, we recursively unify
    $A[\theta]$ with $\epsilon$. If that succeeds, we get a $\theta'$ which
    unifies $A$ and $\epsilon$ and thus forces the constraint to hold, so we
    can again use \LabTirName{metasub/constancy} to solve the constraint.
  \item
    In any other case we simply return $\theta$ and keep the constraint around.
\end{enumerate}

\emph{Remark.} In the case with $x \in \FreeVars(B[\theta])$, it would be also
sound to solve the constraint when the occurrence is not rigid. However, this
way we could lose potential non-$\epsilon$ solutions of $A$.

\subsection{Unification For Strictly Curried Functions}
Although we omit most details of unification, we shall discuss it for curried
functions, as it is essential in the extended elaboration algorithm.  The most
interesting case is when we unify a curried function type with an implicit
function type. In this case, we learn that the domain of the curried function is
non-empty, so we refine the $A$ domain to an extended $(x_0 : A_0) \TCons A_1$
telescope. Since we invent a fresh $A_1$ domain for a curried function type, we
need to add a constancy constraint for it as well.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x_0 : A_0\}\to B') :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1,\,A_1) \leftarrow \freshMeta{\Theta_0}{\Gamma,\,x_0 : A_0}{\Tel}\\
& \quad (\Theta_2,\,\theta_2) \leftarrow
  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A_0[\theta_1])\TCons A_1)\\
& \quad (\Theta_3,\,\theta_3) \leftarrow
    \newConstancy{\Theta_2}{\Gamma[\theta_{12}],\,x_0 : A_0[\theta_{12}]}{x_1}{(A_1[\theta_2])}{(B[\theta_{12}][x\mapsto (x_0 :: x_1)])}\\
& \quad\unify_{\Theta_3|\Gamma[\theta_{123}],\,x_0 : A_0[\theta_{123}]}\,
         (\{x_1 : A_1[\theta_{23}]\}\to B[\theta_{123}][x\mapsto (x_0 :: x_1)])\,B'
\end{alignat*}
We have the symmetric $\unify_{\Theta_0|\Gamma}(\{x_0 : A_0\}\to B')\,(\{x :
\ol{A}\}\to B)$ case the same way as above.

Now, let us assume that $B'$ is not an implicit function type and not
meta-headed. Then, we have the following case, where we solve a telescope domain
to be empty.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,B' :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1) \leftarrow \unify_{\Theta_0|\Gamma}\,A\,\epsilon\\
& \quad \unify_{\Theta_1|\Gamma[\theta_1]}\,(B[\theta_1][x\mapsto[]])\,(B'[\theta_1])
\end{alignat*}
Again, we also have the symmetric case. For $\lambda\,\{x : \ol{A}\}.\,t$ and
$t\,\{u : \ol{A}\}$, unification is structural, and other cases remain the same
as in the the basic elaborator of Section \ref{sec:bidirectional_elaboration}.
